---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->

[![Codecov test coverage](https://codecov.io/gh/Polymerase3/phiper/graph/badge.svg)](https://app.codecov.io/gh/Polymerase3/phiper)
[![R-CMD-check](https://github.com/Polymerase3/phiper/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Polymerase3/phiper/actions/workflows/R-CMD-check.yaml)
[![pkgcheck](https://github.com/Polymerase3/phiper/workflows/pkgcheck/badge.svg)](https://github.com/Polymerase3/phiper/actions?query=workflow%3Apkgcheck)
[![Project Status: Active/Unstable](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)

<!-- badges: end -->

# phiper

The `phiper` package provides a scalable, reproducible toolkit for validating and analyzing PhIP-Seq data in cross-sectional and longitudinal studies.

## Installation

You can install the development version of `phiper` from GitHub with either `pak` or `devtools`:

```{r, eval = FALSE}
# install.packages("pak")
pak::pak("Polymerase3/phiper")

# or, using devtools:
# install.packages("devtools")
devtools::install_github("Polymerase3/phiper")
```

## Usage

For testing, exploration, and minimal examples we provide the `phip_mixture` dataset. It is a simulated PhIP-Seq–like antibody repertoire for two groups (A, B) and two time points (T1, T2), with per-subject, per-peptide presence/absence (`exist`) plus corresponding read counts and fold changes. The dataset is generated from a simple mixture of rare and common peptide prevalences with Poisson-distributed counts and is intended solely for testing and illustrative examples, without any biological interpretation.

### Importing the data

Because the data are relatively large, they are stored in a `.parquet` file. You can obtain the file path with:

```{r, results = "hide", message = FALSE, warning = FALSE}
library(phiper)
library(dplyr)
library(ggplot2)

phip_path <- phip_example_path()
```

You can then import the data with:

```{r, results = "hide", message = FALSE, warning = FALSE}
ps <- phip_convert(
  data_long_path    = phip_path,
  backend           = "duckdb",
  peptide_library   = TRUE,
  subject_id        = "subject_id",
  peptide_id        = "peptide_id",
  sample_id         = "sample_id",
  exist             = "exist",
  timepoint         = "timepoint_factor",
  fold_change       = "fold_change",
  materialise_table = TRUE,
  auto_expand       = FALSE,
  n_cores           = 5
)
```

You can inspect the data:

```{r}
ps$data_long %>%
  head(5) %>%
  collect()
```

You can also inspect the current version of the peptide library maintained by the Vogl lab:

```{r}
get_peptide_library(ps) %>%
  head(5) %>%
  collect()
```

### Plotting peptide counts

You can easily plot peptide counts using the `plot_enrichment_counts()` function:

```{r, message = FALSE, warning = FALSE}
plot_enrichment_counts(
  ps,
  group_cols        = c("group", "time"),
  group_interaction = TRUE,
  interaction_only  = TRUE,
  annotation_size   = 3
)
```

### Alpha diversity

You can also compute and visualize alpha diversity for different groups or subgroups of interest. `phiper` is organized into analysis modules: for each module you typically get a computation function (for example, `compute_alpha_diversity()`) and one or more plotting functions (for example, `plot_alpha_diversity()`). This pattern repeats across the analysis modules.

For alpha diversity specifically:

```{r, message = FALSE, warning = FALSE}
alpha <- compute_alpha_diversity(
  ps,
  group_cols        = c("group", "time"),
  carry_cols        = "subject_id",
  group_interaction = TRUE,
  interaction_only  = TRUE
)

head(alpha)
```

You can then plot the results:

```{r, message = FALSE, warning = FALSE}
plot_alpha_diversity(
  alpha,
  metric    = "richness",
  group_col = "phip_interaction"
) +
  ylim(100, 400) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      vjust = 1,
      hjust = 1
    )
  )
```

## Contributing

Bug reports, feature requests, and questions are welcome. Please open an issue on the GitHub tracker:

- <https://github.com/Polymerase3/phiper/issues>

If you’d like to contribute, please read [CONTRIBUTING.md](CONTRIBUTING.md).
